---
title: "Indirect effects in CRISPR enhancer screens"
author: "Andreas R. Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setupDocument, include=FALSE}
# set output html chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

# suppress dplyrs summarize info
options(dplyr.summarize.inform = FALSE)
```

## Goal
Analyze indirect effects results.
```{r attachPackages, message=FALSE, warning=FALSE}
library(tidyverse)
library(cowplot)
library(DT)
```

***

## Used data
Results from the indirect CRISPR effects and cis differential expression analyses are used for these
analyses.
```{r prepareData}
# load cis and trans positive rates from indirect effects analyses
cis_rates <- read_tsv(snakemake@input$cis_rates, show_col_types = FALSE)
trans_rates <- read_tsv(snakemake@input$trans_rates, show_col_types = FALSE)

# filter cis hit rates to within 1Mb only
cis_rates <- filter(cis_rates, mean_dist < 1e6)

# load direct effect rates per dataset and averaged across datasets
direct_rates_datasets <- read_tsv(snakemake@input$direct_rates_datasets, show_col_types = FALSE)
direct_rates_average  <- read_tsv(snakemake@input$direct_rates_average, show_col_types = FALSE)

# load direct effects models
direct_effect_models <- readRDS(snakemake@input$direct_effect_models)

# load CRISPR data from benchmarking pipeline merged with scores from E-G predictors
merged_training <- read_tsv(snakemake@input$merged_training, show_col_types = FALSE)
merged_heldout  <- read_tsv(snakemake@input$merged_heldout,  show_col_types = FALSE)

# predictor related columns added to CRISPR data by benchmarking pipeline
benchmark_cols <- c("expressed", "PredictionCellType", "pred_elements", "pred_uid", "pred_id",
                    "pred_col", "pred_value", "Prediction")

# extract data on CRISPR E-G pairs only from merged data
crispr_training <- merged_training %>% 
  filter(pred_uid == "baseline.distToTSS") %>% 
  mutate(dist_to_tss = pred_value) %>% 
  select(-any_of(benchmark_cols))

crispr_heldout <- merged_heldout %>% 
  filter(pred_uid == "baseline.distToTSS") %>% 
  mutate(dist_to_tss = pred_value) %>% 
  select(-any_of(benchmark_cols))

# load file used to convert dataset ids in CRISPR data files to ids in indirect effects analyses
datasets <- read_tsv(snakemake@input$dataset_ids, show_col_types = FALSE)
datasets <- rename(datasets, indirect_effects_dataset = dataset, Dataset = dataset_crispr_files)

# combine all CRISPR data into one table and add dataset ids for indirect effects analyses
crispr <- bind_rows(training = crispr_training, heldout = crispr_heldout, .id = "dataset_type") %>% 
  left_join(datasets, by = "Dataset") %>% 
  mutate(dataset_plots = factor(dataset_plots, levels = unique(datasets$dataset_plots)))
```

***

## Expected indirect hits
Calculate the number of expected indirect effects in cis E-G analyses and the expected proportion of
cis hits to be indirect effects for negative, positive or any effect size direction.
```{r}
# count the number of cis pairs and hits per dataset across all distances and convert to long format
effect_rates <- cis_rates %>% 
  group_by(dataset) %>% 
  summarize(total_pairs = sum(total_pairs),
            significant_pairs = sum(significant_pairs),
            negative_pairs = sum(negative_pairs),
            positive_pairs = sum(positive_pairs),
            .groups = "drop") %>% 
  pivot_longer(cols = -c(dataset, total_pairs), names_to = "effect_direction",
               values_to = "significant_cis_pairs") %>% 
  select(dataset, effect_direction, tested_cis_pairs = total_pairs, significant_cis_pairs) %>% 
  mutate(effect_direction = sub("_pairs", "", effect_direction)) %>% 
  mutate(effect_direction = if_else(effect_direction == "significant", true = "any",
                                    false = effect_direction))

# convert indirect (trans) hits to long format and add to cis rates
effect_rates <- trans_rates %>% 
  select(dataset, tested_indirect_pairs = total_pairs, any = significant_pairs,
         negative = negative_pairs, positive = positive_pairs) %>% 
  pivot_longer(cols = -c(dataset, tested_indirect_pairs), names_to = "effect_direction",
               values_to = "significant_indirect_pairs") %>% 
  left_join(effect_rates, ., by = c("dataset", "effect_direction"))

# convert indirect (trans) hit rates to long format and add to cis rates
effect_rates <- trans_rates %>% 
  select(dataset, any = positive_rate_significant, negative = positive_rate_negative,
         positive = positive_rate_positive) %>% 
  pivot_longer(cols = -dataset, names_to = "effect_direction",
               values_to = "indirect_effects_rate") %>% 
  left_join(effect_rates, ., by = c("dataset", "effect_direction"))

# compute number and percentage of expected indirect effects
effect_rates <- effect_rates %>% 
  mutate(expected_indirect_cis_pairs = round(tested_cis_pairs * indirect_effects_rate)) %>% 
  mutate(percent_indirect_cis_pairs = expected_indirect_cis_pairs / significant_cis_pairs)
```

Compute how many of the final, filtered perturbation-gene pairs would can be expected to be from
indirect effects, based on the estimated percentage of indirect effects among all cis positives.
```{r}
# get filtered final number of significant pairs
final_significant_pairs <- crispr %>%
  filter(ValidConnection == TRUE) %>% 
  group_by(indirect_effects_dataset) %>%
  summarize(any = sum(Significant == TRUE),
         negative = sum(Significant == TRUE & EffectSize <= 0),
         positive = sum(Significant == TRUE & EffectSize > 0)) %>% 
  pivot_longer(cols = -indirect_effects_dataset, names_to = "effect_direction",
               values_to = "filtered_significant_cis_pairs")
  
# add to table with calculated effect rates
effect_rates <- left_join(effect_rates, final_significant_pairs,
                          by = c("dataset" = "indirect_effects_dataset", "effect_direction"))

# compute expected number of filtered pairs to be due to indirect effects based on
effect_rates <- effect_rates %>% 
  mutate(expected_indirect_final_cis_pairs = round(filtered_significant_cis_pairs * percent_indirect_cis_pairs))

# save effect rates to output file
write_tsv(effect_rates, "results/manuscript/tables/crispr_indirect_effects.tsv")
```

```{r}
datatable(
  effect_rates,
  extensions = c("FixedColumns", "Buttons"),
  options = list(
    pageLength = 20,
    dom = 'Bfrtip',
    scrollX = TRUE,
    fixedColumns = list(leftColumns = 3),
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')),
  autoHideNavigation = TRUE) %>%
  formatStyle(c(2, 4, 7, 9), `border-right` = "solid 1px")
```

***

## Direct effects as function of distance
```{r, fig.width=11, fig.height=4}
# add datasets to use in plots to direct rates
direct_rates_datasets <- datasets %>% 
  select(dataset = indirect_effects_dataset, dataset_plots) %>% 
  left_join(direct_rates_datasets, ., by = "dataset")

# get direct effect rates for enhancer (negative) effects
direct_rates_avg_enh <- filter(direct_rates_average, type == "direct_rate_negative")
direct_rates_datasets_enh <- filter(direct_rates_datasets, type == "direct_rate_negative")
  
# get model coefficients for enhancer (negative) effects
model <- direct_effect_models[["direct_rate_negative"]]
a <- exp(coef(model)[1])
b <- coef(model)[2]
  
# get maximum direct rate as y axis maximum
ylim <- max(direct_rates_datasets_enh$direct_rate)

# plot model fit vs average direct rate per dataset
p1 <- ggplot(direct_rates_avg_enh, aes(x = dist_to_tss, y = direct_rate)) +
  geom_point(color = "steelblue") +
  stat_function(fun = function(x) a * x^b, color = "black") +
  labs(title = "Direct hit rate (avg. across datasets)", x = "Distance to TSS (bp)",
       y = "Direct hit rate") +
  scale_x_continuous(labels = scales::label_scientific(digits = 1)) +
  scale_y_continuous(limits = c(NA, ylim)) +
  theme_classic() +
  theme(text = element_text(size = 13))
  
# plot model fit vs. direct rate per individual dataset
p2 <- ggplot(direct_rates_datasets_enh,
             aes(x = dist_to_tss, y = direct_rate, color = dataset_plots)) +
  geom_point() +
  stat_function(fun = function(x) a * x^b, color = "black") +
  labs(title = "Direct hit rate (per dataset)", x = "Distance to TSS (bp)", y = "Direct hit rate",
       color = "Dataset") +
  scale_x_continuous(labels = scales::label_scientific(digits = 1)) +
  scale_y_continuous(limits = c(NA, ylim)) +
  theme_classic() +
  theme(text = element_text(size = 13))

# arrange into one plot
plot_grid(p1, p2, nrow = 1, scale = 0.95, rel_widths = c(0.62, 1))
```

```{r, fig.width=11, fig.height=4}
# convert plots to log space
p1 <- p1 +
  scale_x_log10() +
  scale_y_log10(limits = c(NA, ylim)) +
  annotation_logticks()

p2 <- p2 +
  scale_x_log10() +
  scale_y_log10(limits = c(NA, ylim)) +
  annotation_logticks()
    

# arrange into one plot
plot_grid(p1, p2, nrow = 1, scale = 0.95, rel_widths = c(0.62, 1))
```


***

## Indirect effects as function of distance
Using the model for direct effects as function of distance to TSS and the estimated indirect effects
rate the probability of hits being the result of direct hits as function of their distance to TSS
was calculated.

### Imputing indirect effects rates
For some CRISPR datasets in both the training and held-out data, no empirical indirect effect rates
was calculated. To calculate the probability of being direct effects, the average indirect effects
rate across all datasets for which indirect rates were computed was used.

Datasets for which the indirect effects rate was imputed:
```{r}
crispr %>% 
  filter(imputed_indirect_rate == TRUE) %>% 
  select(`Dataset type` = dataset_type, Dataset = indirect_effects_dataset, Regulated) %>% 
  group_by(`Dataset type`, Dataset) %>% 
  summarize(pairs = n(), positives = sum(Regulated == TRUE), .groups = "drop") %>% 
  datatable()
```

### Probability of direct effects
Plot the probability of E-G pairs being a direct vs. indirect effect as a function of distance to 
TSS.
```{r, fig.height=4, fig.width=6.2}
# add effects direction label for each CRISPR E-G pair
crispr <- crispr %>% 
  mutate(effect_direction = if_else(EffectSize <= 0, true = "Negative", false = "Positive"))

# plot direct vs. indirect effect probability as function of distance to TSS
ggplot(crispr, aes(x = dist_to_tss / 1e3, y = direct_vs_indirect_negative,
                            color = dataset_plots)) +
  geom_line() +
  geom_point(data = filter(crispr, Regulated == TRUE), size = 1.75) +
  geom_hline(yintercept = c(0.5, 0.9), linetype = "dashed", color = "black", linewidth = 0.25) +
  labs(title = "Probability of direct CRISPR effects", x = "Distance to TSS (kb)",
       y = "Probability direct vs. indirect effect", color = "Dataset",
       shape = "Significant\neffect direction") +
  scale_shape_manual(values = c(Negative = 19, Positive = 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 1050)) +
  theme_classic()

# save plot to pdf
ggsave(file = "results/manuscript/plots/crispr_direct_vs_indirect_probability.pdf", height = 4,
       width = 6.2)
```

***

## Likely direct and indirect effects
Using the model for direct effects as function of distance to TSS and the estimated indirect effects
rate calculate at which distance CRISPR hits have >90% and <10% probability to be direct effects.
```{r}
# function to predict the direct hit rate for a given distance
predict_direct_hit_rate <- function(models, new_data, type) {
  
  # get model for specified type
  fit <- models[[type]]
  
  # predict direct hit rate for provided distances and exponentiate to get back to original scale
  predicted_log_direct_rate <- predict(fit, newdata = new_data)
  predicted_direct_rate <- exp(predicted_log_direct_rate)
  
  # cap predicted values at 1, because the rate can't be higher than that
  predicted_direct_rate <- pmin(predicted_direct_rate, 1)
  
  return(predicted_direct_rate)
  
}

# function to find distance to TSS where CRISPR pairs where probability of being direct effects is
# greater than a given probability.
calculate_dist_direct_gt_prob <- function(dataset, direct_effect_models, dist_range = c(1, 1e6),
                                          indirect_rates, prob) {
  
  # create distance to TSS range in 1bp increments
  dist_range <- tibble(
    dataset = dataset, 
    dist_to_tss = seq(from = dist_range[[1]], to = dist_range[[2]], by = 1)
  )
  
  # predict direct effect rates based on distance to TSS
  dist_range <- dist_range %>% 
    mutate(direct_hit_rate = predict_direct_hit_rate(direct_effect_models, new_data = dist_range,
                                                     type = "direct_rate_negative"))
  
  # add indirect effects rate for each dataset
  dist_range <- indirect_rates %>% 
    select(dataset, indirect_hit_rate = positive_rate_negative) %>% 
    left_join(dist_range, ., by = "dataset")
  
  # calculate probability direct vs. indirect hit
  dist_range <- dist_range %>% 
    mutate(direct_vs_indirect =  direct_hit_rate / (direct_hit_rate + indirect_hit_rate))
  
  # get distance at which probability of being a direct effect is greater than 'prob'
  dist_prob <- max(dist_range[dist_range$direct_vs_indirect > prob, "dist_to_tss"])
  dist_prob <- tibble(dataset, !!paste("distance prob. direct vs indirect >", prob) := dist_prob)
  
  return(dist_prob)
  
}
```

```{r}
# get all CRISPR datasets in training and heldout data
crispr_datasets <- as.character(unique(crispr$dataset_plots))
names(crispr_datasets) <- crispr_datasets

# extract empirical or imputed indirect rates for them
indirect_rates <- crispr %>% 
  select(dataset = dataset_plots,
         positive_rate_significant = indirect_rate_all,
         positive_rate_negative = indirect_rate_negative,
         positive_rate_positive = indirect_rate_positive) %>% 
  distinct()

# compute distance at which probability of being a direct effect vs indirect is > 90% for all 
# datasets
dist_direct_probs <- lapply(crispr_datasets, FUN = calculate_dist_direct_gt_prob,
                            direct_effect_models = direct_effect_models,
                            indirect_rates = indirect_rates, prob = 0.9)

# combine into one table
dist_direct_probs <- bind_rows(dist_direct_probs, .id = "dataset")
```

Distance to TSS at which probability of CRISPR E-G pairs being direct vs. indirect is grated than
0.9:
```{r}
datatable(dist_direct_probs)
```

### Enrichment for negative effects among likely direct pairs
```{r}
# annotate results of direct vs indirect effect probability is >0.9
crispr <- crispr %>% 
  mutate(p90 = if_else(direct_vs_indirect_negative > 0.9, true = "Prob. direct > 0.9",
                       false = "Prob. direct <= 0.9"))

# count number of significant pairs with positive and negative effects for pairs with direct vs
# indirect effect probability greater or lower than 0.9
pairs_p90 <- crispr %>% 
  group_by(dataset_plots, p90) %>% 
  summarize(significant = sum(Significant == TRUE),
            negative = sum(Significant == TRUE & EffectSize <= 0),
            positive = sum(Significant == TRUE & EffectSize > 0),
            prop_negative = negative / significant,
            prop_positive = positive / significant,
            .groups = "drop")
```

```{r, fig.width=12, fig.height=5}
# transform to long format for plotting
pairs_p90_plot <- pairs_p90 %>% 
  select(dataset_plots, p90, prop_negative, prop_positive) %>% 
  pivot_longer(cols = c(prop_negative, prop_positive), names_to = "Effect direction",
               values_to = "% significant pairs") %>% 
  mutate(`Effect direction` = str_to_title(sub("prop_", "", `Effect direction`)))

# plot the proportion of negative vs positives based on probability to be a direct effect
ggplot(pairs_p90_plot, aes(x = fct_rev(p90), y = `% significant pairs`, fill = `Effect direction`)) +
  facet_wrap(~ dataset_plots, nrow = 1) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linewidth = 0.25) +
  labs(title = "Negative vs. postive effects") +
  scale_fill_manual(values = c(Negative = "steelblue", Positive = "firebrick3")) +
  scale_y_continuous(labels = scales::percent) +
  theme_classic() +
  theme(axis.title.x = element_blank(), strip.background = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Effect sizes vs. probability of direct effects
```{r, fig.width=7, fig.height=5}
# get all significant hits
crispr_hits <- filter(crispr, Significant == TRUE)

# plot CRISPR effect size of likely direct vs. likely indirect effects
ggplot(crispr_hits, aes(x = dataset_plots, y = EffectSize, color = fct_rev(p90),
                                 group = interaction(dataset_plots, fct_rev(p90)))) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
  geom_point(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.3)) +
  geom_boxplot(color = "black", fill = NA, outlier.shape = NA) +
  labs(title = "CRISPR effect size likely direct vs. indirect effects",
       y = "CRISPR effect size\n(percent change)", color  = "Probability\ndirect vs indirect") +
  scale_color_manual(values = c(`Prob. direct > 0.9` = "firebrick3", `Prob. direct <= 0.9` = "steelblue")) +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, fig.width=7, fig.height=5}
# convert effect size from percent change to log2 fold-change
crispr_hits <- crispr_hits %>% 
  mutate(log2_fc_effect_size = log2(1 + EffectSize))

# plot CRISPR effect size of likely direct vs. likely indirect effects
ggplot(crispr_hits, aes(x = dataset_plots, y = log2_fc_effect_size, color = fct_rev(p90),
                        group = interaction(dataset_plots, fct_rev(p90)))) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5) +
  geom_point(alpha = 0.5, position = position_jitterdodge(jitter.width = 0.3)) +
  geom_boxplot(color = "black", fill = NA, outlier.shape = NA) +
  labs(title = "CRISPR effect size likely direct vs. indirect effects",
       y = "CRISPR effect size\n(log2 fold change)", color  = "Probability\ndirect vs indirect") +
  scale_color_manual(values = c(`Prob. direct > 0.9` = "firebrick3", `Prob. direct <= 0.9` = "steelblue")) +
  theme_classic() +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

***

## Probability of direct effects of long-range interactions
Calculate the average probability of direct effects for interactions below and beyond 100kb.
```{r}
# probability of direct / indirect effects per dataset
crispr %>% 
  mutate(distance_category = if_else(dist_to_tss > 1e5, true = ">100kb", false = "<=100kb")) %>%
  group_by(dataset_type, Dataset, distance_category) %>% 
  summarize(avg_prob_direct_effect = mean(direct_vs_indirect_negative), .groups = "drop") %>% 
  mutate(avg_prob_indirect_effect = 1 - avg_prob_direct_effect) %>% 
  datatable()
```

```{r}
# probability of direct / indirect effects in combined datasets
crispr %>% 
  mutate(distance_category = if_else(dist_to_tss > 1e5, true = ">100kb", false = "<=100kb")) %>%
  group_by(dataset_type, distance_category) %>% 
  summarize(avg_prob_direct_effect = mean(direct_vs_indirect_negative), .groups = "drop") %>% 
  mutate(avg_prob_indirect_effect = 1 - avg_prob_direct_effect) %>% 
  datatable()
```

***

## Session information
This document was produced using following packages:
```{r sessionInfo}
sessionInfo()
```
