---
title: "Extended Data Fig. 2 - CRISPR benchmarking"
author: "Andreas R. Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setupDocument, include=FALSE}
# set output html chunk options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

# set random seed
set.seed(snakemake@params$seed)
```

## Goal
Create plots for CRISPR benchmarking Extended Data Figure 2.

```{r attachPackages, message=FALSE, warning=FALSE}
# attach required packages
library(tidyverse)
library(rtracklayer)
library(cowplot)
library(ROCR)
library(caTools)

# required functions
proj_dir <- snakemake@config$proj_dir
source(file.path(proj_dir, "CRISPR_benchmarks/workflow/scripts/crisprComparisonLoadInputData.R"))
source(file.path(proj_dir, "CRISPR_benchmarks/workflow/scripts/crisprComparisonPlotFunctions.R"))
source(file.path(proj_dir, "CRISPR_benchmarks/workflow/scripts/crisprComparisonBootstrapFunctions.R"))
source("workflow/scripts/random_gene_pred_functions.R")
```

***

## DHS Baseline predictors

Plot Precision-Recall curves for all baseline predictors based on DHS sites.
```{r}
# load TSS annotations
tss_annot <- read_tsv(snakemake@input$tss_annot, show_col_types = FALSE, col_select = 1:6)

# create GRanges object with TSS annotations
tss_annot <- makeGRangesFromDataFrame(tss_annot, seqnames.field = "#chr", keep.extra.columns = TRUE,
                                      starts.in.df.are.0based = TRUE)

# load list of expressed genes in K562
expr_genes <- fread(snakemake@input$expr_genes)

# load pred_config file
pred_config_bl <- importPredConfig(snakemake@input$pred_config, expr = TRUE,
                                   include_col = "crispr_baseline_figure")

# load merged data for DHS baseline predictors
merged_dhs <- fread(snakemake@input$merged_baseline_dhs)

# process merged data for benchmarking analyses, including filtering for ValidConnection == TRUE
merged_dhs <- processMergedData(merged_dhs, pred_config = pred_config_bl,
                                filter_valid_connections = TRUE, include_missing_predictions = TRUE)
```

```{r}
# compute precision-recall tables
pr_dhs <- calcPRCurves(merged_dhs, pred_config = pred_config_bl, pos_col = "Regulated")

# calculate number and percentage of experimental true positives in the CRISPR datasets
n_pos_dhs <- calcNPos(merged_dhs, pos_col = "Regulated")
pct_pos_dhs <- calcPctPos(merged_dhs, pos_col = "Regulated")

# get number of tested enhancer gene pairs in merged data and create title for PR plot
n_pairs_dhs <- n_distinct(merged_dhs$name)
pr_title_dhs <- paste0("DHS baseline predictors", " (", scales::label_comma()(n_pairs_dhs), " E-G pairs)")

# get colors for predictors in plots
pred_colors_dhs <- deframe(select(pred_config_bl, pred_name_long, color))
```

```{r PRCFunction}
# create PRC plots with sets of predictors highlighted at a time
plotPRCsets <- function(pr_df, pred_config, sets, n_pos, pct_pos, plot_name, pred_colors,
                        min_sensitivity = 0.7, line_width = 1.2, point_size = 3.5, text_size = 13) {
  
  # get names of sets for plot titles
  titles <- names(sets)
  
  # create plots with highlighted sets of predictors
  mapply(sets, titles, FUN = plotPRCSset,
         MoreArgs = list(pr_df = pr_df, pred_config = pred_config, n_pos = n_pos, pct_pos = pct_pos,
                        min_sensitivity = min_sensitivity, line_width = line_width,
                        point_size = point_size, text_size = text_size, pred_colors = pred_colors),
         SIMPLIFY = FALSE)
  
}

# function to plot PRCs and highlight one set of predictors
plotPRCSset <- function(pr_df, pred_config, set, n_pos, pct_pos, plot_name, min_sensitivity,
                        line_width, point_size, text_size, pred_colors) {
  
  # get colors for set of predictors and set others to NA
  pred_colors[!names(pred_colors) %in% set] <- NA_character_
  
  # create PRC plot
  makePRCurvePlot(pr_df = pr_df, pred_config = pred_config, n_pos = n_pos, pct_pos = pct_pos,
                  min_sensitivity = min_sensitivity, plot_name = plot_name, line_width = line_width,
                  point_size = point_size, text_size = text_size, colors = pred_colors,
                  plot_thresholds = FALSE) +
    geom_vline(xintercept = 0.7, linetype = "dashed", color = "black")
  
}
```

```{r}
# sets of predictors to highlight
pred_sets_dhs <- list(
  "Distance" = c(
    "ENCODE-rE2G_Extended",
    "In element (DHS) & distance to TSS",
    "In element (DHS) & distance to gene",
    "In element (DHS) & closest gene",
    "In element (DHS) & closest expr. TSS",
    "In element (DHS) & closest TSS",
    "In element (DHS) & closest expr. gene",
    "In element (DHS) & within 100kb of TSS",
    "Distance to TSS"
    ),
  "Activity x Contact" = c(
    "ENCODE-rE2G_Extended",
    "ABC_A=DNase x H3K27ac, C=ENCODE Hi-C",
    "In element (DHS) & DNase RPM x distance",
    "In element (DHS) & DNase RPM x distance (norm.)",
    "In element (DHS) & H3K27ac RPM x distance",
    "In element (DHS) & H3K27ac RPM x distance (norm.)",
    "Distance to TSS"
    )
  )

# get pred_uid for these
plot_preds <- unlist(pred_sets_dhs)
plot_pred_uids <- pred_config_bl %>% 
  filter(pred_name_long %in% plot_preds) %>% 
  pull(pred_uid)

# filter pr object and colors only for predictors to include in plots
pr_dhs <- filter(pr_dhs, pred_uid %in% plot_pred_uids)
pred_colors_dhs <- pred_colors_dhs[names(pred_colors_dhs) %in% plot_preds]

# manually set color for Distance to TSS, since gray is used to prevent over-plotting
pred_colors_dhs[["Distance to TSS"]] <- "#FFA600"

# create set plots
dhs_set_plots <- plotPRCsets(pr_df = pr_dhs, pred_config = pred_config_bl, sets = pred_sets_dhs,
                             n_pos = n_pos_dhs, pct_pos = pct_pos_dhs,
                             pred_colors = pred_colors_dhs, min_sensitivity = 0.7,
                             line_width = 0.75, point_size = 3, text_size = 12)
```

```{r}
# create plot with all predictors for whole dataset
dhs_all_prc <- makePRCurvePlot(pr_df = pr_dhs, pred_config = pred_config_bl, n_pos = n_pos_dhs,
                               pct_pos = pct_pos_dhs, min_sensitivity = 0.7, line_width = 1,
                               point_size = 3, text_size = 12, colors = pred_colors_dhs)

# extract legend to use for all set plots
dhs_legend <- get_legend(
  dhs_all_prc + 
    labs(color = "Predictor:") +
    theme(legend.key.size = unit(1, "lines"))
  )

# create title for arranged plots
dhs_title <- ggdraw() + 
  draw_label(pr_title_dhs, fontface = 'bold', size = 15, x = 0, hjust = 0)

# arrange PRCs into grid
dhs_prcs <- plot_grid(
  dhs_set_plots[[1]] + theme_classic() + theme(legend.position = "none"),
  dhs_set_plots[[2]] + theme_classic() + theme(legend.position = "none"),
  dhs_legend,
  nrow = 1,
  scale = 0.9
)

# add legend to create figure panel
dhs_baseline_panel <- plot_grid(
  dhs_title + theme(plot.margin = margin(0, 0, 0, 22)),
  dhs_prcs,
  nrow = 2,
  rel_heights = c(0.05, 1)
)
```

## ABC Baseline predictors

Plot Precision-Recall curves for all baseline predictors based on ABC sites.
```{r}
# load merged data for published predictors
merged_abc <- fread(snakemake@input$merged_baseline_abc)

# process merged data for benchmarking analyses, including filtering for ValidConnection == TRUE
merged_abc <- processMergedData(merged_abc, pred_config = pred_config_bl,
                                filter_valid_connections = TRUE, include_missing_predictions = TRUE)
```

```{r}
# compute precision-recall tables
pr_abc <- calcPRCurves(merged_abc, pred_config = pred_config_bl, pos_col = "Regulated")

# calculate number and percentage of experimental true positives in the CRISPR datasets
n_pos_abc <- calcNPos(merged_abc, pos_col = "Regulated")
pct_pos_abc <- calcPctPos(merged_abc, pos_col = "Regulated")

# get number of tested enhancer gene pairs in merged data and create title for PR plot
n_pairs_abc <- n_distinct(merged_abc$name)
pr_title_abc <- paste0("ABC baseline predictors", " (", scales::label_comma()(n_pairs_abc), " E-G pairs)")

# get colors for predictors in plots
pred_colors_abc <- deframe(select(pred_config_bl, pred_name_long, color))
```

```{r}
# sets of predictors to highlight
pred_sets_abc <- list(
  "Distance" = c(
    "ENCODE-rE2G_Extended",
    "In element (ABC) & distance to TSS",
    "In element (ABC) & distance to gene",
    "In element (ABC) & closest gene",
    "In element (ABC) & closest expr. TSS",
    "In element (ABC) & closest TSS",
    "In element (ABC) & closest expr. gene",
    "In element (ABC) & within 100kb of TSS",
    "Distance to TSS"
    ),
  "Activity x Contact" = c(
    "ENCODE-rE2G_Extended",
    "ABC_A=DNase x H3K27ac, C=ENCODE Hi-C",
    "In element (ABC) & DNase RPM x distance",
    "In element (ABC) & DNase RPM x distance (norm.)",
    "In element (ABC) & H3K27ac RPM x distance",
    "In element (ABC) & H3K27ac RPM x distance (norm.)",
    "Distance to TSS"
    )
  )

# get pred_uid for these
plot_preds <- unlist(pred_sets_abc)
plot_pred_uids <- pred_config_bl %>% 
  filter(pred_name_long %in% plot_preds) %>% 
  pull(pred_uid)

# filter pr object only for predictors to include in plots
pr_abc <- filter(pr_abc, pred_uid %in% plot_pred_uids)

# only retain colors for predictors to show
pred_colors_abc <- pred_colors_abc[names(pred_colors_abc) %in% plot_preds]

# manually set color for Distance to TSS, since gray is used to prevent over-plotting
pred_colors_abc[["Distance to TSS"]] <- "#FFA600"

# create set plots
abc_set_plots <- plotPRCsets(pr_df = pr_abc, pred_config = pred_config_bl, sets = pred_sets_abc,
                             n_pos = n_pos_abc, pct_pos = pct_pos_abc,
                             pred_colors = pred_colors_abc, min_sensitivity = 0.7,
                             line_width = 0.75, point_size = 3, text_size = 12)
```

```{r}
# create plot with all predictors
abc_all_prc <- makePRCurvePlot(pr_df = pr_abc, pred_config = pred_config_bl, n_pos = n_pos_abc,
                               pct_pos = pct_pos_abc, min_sensitivity = 0.7, line_width = 1,
                               point_size = 3, text_size = 12, colors = pred_colors_abc)

# extract legend to use for all set plots
abc_legend <- get_legend(
  abc_all_prc + 
    labs(color = "Predictor:") +
    theme(legend.key.size = unit(1, "lines"))
  )

# create title for arranged plots
abc_title <- ggdraw() + 
  draw_label(pr_title_abc, fontface = 'bold', size = 15, x = 0, hjust = 0)

# arrange PRCs into grid
abc_prcs <- plot_grid(
  abc_set_plots[[1]] + theme_classic() + theme(legend.position = "none"),
  abc_set_plots[[2]] + theme_classic() + theme(legend.position = "none"),
  abc_legend,
  nrow = 1,
  scale = 0.9
)

# add legend to create figure panel
abc_baseline_panel <- plot_grid(
  abc_title + theme(plot.margin = margin(0, 0, 0, 22)),
  abc_prcs,
  nrow = 2,
  rel_heights = c(0.05, 1)
)
```

## Published predictors
Make Precision-Recall and performance plots for other published semi-supervised predictors.

### Precision-Recall curves
```{r}
# load merged data for published predictors
merged_pub <- fread(snakemake@input$merged_published_preds)

# load pred_config file for published predictions
pred_config_pub <- importPredConfig(snakemake@input$pred_config, expr = TRUE,
                                    include_col = "crispr_published_pred_figure")

# process merged data for benchmarking analyses, including filtering for ValidConnection == TRUE
merged_pub <- processMergedData(merged_pub, pred_config = pred_config_pub,
                                filter_valid_connections = TRUE, include_missing_predictions = TRUE)
```

```{r}
# compute performance for published predictive models
pr_pub <- calcPRCurves(merged_pub, pred_config = pred_config_pub, pos_col = "Regulated")
perf_summary_pub <- makePRSummaryTable(pr_pub, pred_config = pred_config_pub, min_sensitivity = 0.7)

# define baseline predictors to show
preds_pub_bl <- c("Distance to TSS", "Closest expressed TSS", "Closest expressed gene",
                  "Within 100kb of expr. TSS")

# extract predictor names ordered by AUPRC performance
preds_pub_quant <- perf_summary_pub %>% 
  arrange(desc(AUPRC)) %>% 
  filter(!pred_name_long %in% preds_pub_bl) %>% 
  pull(pred_name_long)

# get booleas published predictors
preds_pub_bool <- merged_pub %>%
  filter(pred_id != "baseline", !pred_name_long %in% preds_pub_quant) %>%
  pull(pred_name_long) %>%
  unique()

# get colors for predictors in plots ordered by AUPRC (for quantitative predictors)
pred_colors_pub <- deframe(select(pred_config_pub, pred_name_long, color))
pred_colors_pub <- pred_colors_pub[c(preds_pub_quant, preds_pub_bool, preds_pub_bl)]

# calculate number and percentage of experimental true positives in the CRISPR datasets
n_pos_pub <- calcNPos(merged_pub, pos_col = "Regulated")
pct_pos_pub <- calcPctPos(merged_pub, pos_col = "Regulated")

# get number of tested enhancer gene pairs in merged data and create title for PR plot
pr_title_pub <- paste0("Published Predictors")

# create PR curve plot
prc_pub <- makePRCurvePlot(pr_pub, n_pos = n_pos_pub, pct_pos = pct_pos_pub,
                           plot_name = pr_title_pub, pred_config = pred_config_pub,
                           min_sensitivity = 0.7, line_width = 0.75, point_size = 3,
                           colors = pred_colors_pub, plot_thresholds = FALSE)

# add custom plotting theme
prc_pub <- prc_pub +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "black") +
  theme_classic() +
  theme(text = element_text(size = 12), legend.key.size = unit(0.5, "lines"),
        plot.title = element_text(size = 12))
```

### Performance summary
```{r}
# get thresholds @ 70% recall
thresholds_pub <- deframe(select(perf_summary_pub, pred_uid, alpha_at_min_sensitivity))

# convert merged data to wide format for bootstrapping
merged_pub_bs <- convertMergedForBootstrap(merged_pub, pred_config = pred_config_pub)

# run bootstraps to get 95% intervals for AUPRC and precision at threshold
auprc_pub <- bootstrapPerformanceIntervals(merged_pub_bs, metric = "auprc", conf = 0.95,
                                           R = snakemake@params$bootstrap_iterations,
                                           ci_type = "perc", ncpus = 2)
precision_pub <- bootstrapPerformanceIntervals(merged_pub_bs, metric = "precision", 
                                               thresholds = thresholds_pub, conf = 0.95,
                                               R = snakemake@params$bootstrap_iterations,
                                               ci_type = "perc", ncpus = 2)

# add pretty predictor names
auprc_pub <- left_join(auprc_pub, select(pred_config_pub, pred_uid, pred_name_long),
                       by = c("id" = "pred_uid"))
precision_pub <- left_join(precision_pub, select(pred_config_pub, pred_uid, pred_name_long),
                       by = c("id" = "pred_uid"))

# make AUPRC barplot for published predictors
p1 <- ggplot(auprc_pub, aes(y = fct_reorder(pred_name_long, .x = full), x = full,
                            fill = pred_name_long)) +
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(xmin = lower, xmax = upper), position = position_dodge(0.9), width = 0.4,
                color = "black") +
  labs(title = "Area under PRC", x = "AUPRC", fill = "Predictor") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_fill_manual(values = pred_colors_pub) +
  theme_classic() +
  theme(axis.title.y = element_blank(), legend.position = "none")

# make precision @ threshold barplot for published predictors
p2 <- ggplot(precision_pub, aes(y = fct_reorder(pred_name_long, .x = full), x = full,
                                fill = pred_name_long)) +
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(xmin = lower, xmax = upper), position = position_dodge(0.9), width = 0.4,
                color = "black") +
  labs(title = "Precision at 70% recall", x = "Precision", fill = "Predictor") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_fill_manual(values = pred_colors_pub) +
  theme_classic() +
  theme(axis.title.y = element_blank(), legend.position = "none")

# arrange plots
performance_pub <- plot_grid(p1, p2, rel_widths = c(1, 1))
```

***

## Correlation between predictors
```{r}
# load merged data for the combined CRISPR dataset
merged_main <- fread(snakemake@input$merged_main_preds)

# pred config for main comparisons
pred_config_main <- importPredConfig(snakemake@input$pred_config, expr = TRUE,
                                     include_col = "crispr_main_benchmarks")

# process merged data for benchmarking analyses, including filtering for ValidConnection == TRUE
merged_main <- processMergedData(merged_main, pred_config = pred_config_main,
                                 filter_valid_connections = TRUE,
                                 include_missing_predictions = TRUE)

# list of predictors to show in supplementary plots
plot_preds <- grep("baseline", x = pred_config_main$pred_uid, invert = TRUE, value = TRUE)
plot_preds <- c(plot_preds, "baseline.distToTSS")

# subset merged data for predictors that should be plotted
merged_plot <- filter(merged_main, pred_uid %in% plot_preds)
```

```{r}
# extract predictor scores and transform to wide format
pred_scores <- merged_plot %>% 
  select(name, pred_name_long, pred_value) %>% 
  pivot_wider(names_from = pred_name_long, values_from = pred_value)

# compute correlation coefficients between predictors
cor_matrix <- cor(select(pred_scores, -name), method = "spearman")

# title for correlation matrix plot
title = "Correlation of predictor scores"
  
# plot correlation matrix for scores of different predictors
corr_plot <- ggcorrplot(cor_matrix, hc.order = TRUE, type = "lower", lab = TRUE, title = title,
                        lab_size = 2, tl.cex = 8)
```

***

## Correlation of predictor scores with CRISPRi effect sizes
```{r}
# helper function to compute correlation coefficient including confidence intervals and p-value
compute_corr <- function(x) {
  cor_results <- cor.test(x$pred_value, x$EffectSize)
  cor_coeff <- cor_results$estimate
  pvalue <- cor_results$p.value
  conf_int <- as.numeric(cor_results$conf.int)
  data.frame(cor_coeff, pvalue, conf_int_upper = conf_int[[1]], conf_int_lower = conf_int[[2]],
             row.names = NULL)
}
```

```{r}
# transform distance to TSS to 1/distance to TSS for correlation analysis
merged_plot <- merged_plot %>% 
  mutate(pred_value = if_else(pred_uid == "baseline.distToTSS",
                              true = log10(1/pred_value),
                              false = pred_value)) %>% 
  mutate(pred_name_long = if_else(pred_uid == "baseline.distToTSS",
                                  true = "log10(1/Distance to TSS)",
                                  false = pred_name_long))

# compute pearson correlation coefficient of predictor scores vs CRISPR effect size
pearson_cor <- merged_plot %>% 
  split(f = merged_plot$pred_name_long) %>% 
  lapply(FUN = compute_corr) %>% 
  bind_rows(.id = "pred_name_long") %>% 
  mutate(pred_name_long = fct_reorder(pred_name_long, cor_coeff, .desc = TRUE))

# get colors for correlation predictors and adapt to new distance name
pred_colors_cor <- pred_config_main %>% 
  filter(pred_uid %in% plot_preds) %>% 
  select(pred_name_long, color) %>% 
  mutate(pred_name_long = sub("Distance to TSS", "log10(1/Distance to TSS)", pred_name_long)) %>% 
  deframe()

# plot pearson's correlation coefficients
crispr_corr <- ggplot(pearson_cor, aes(x = pred_name_long, y = cor_coeff, color = pred_name_long,
                                fill = pred_name_long)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = conf_int_lower, ymax = conf_int_upper), color = "black", width = 0.2) +
  labs(y = expression(paste("Pearson's ", rho)),
       title = "Predictor scores vs. CRISPR effect size", color = "Predictor", fill = "Predictor") +
  coord_flip() +
  scale_color_manual(values = pred_colors_cor) +
  scale_fill_manual(values = pred_colors_cor) +
  theme_classic() +
  theme(axis.title.y = element_blank(), plot.title = element_text(size = 15, hjust = 1.25),
        legend.position = "none")
```

***

## Assemble supplementary figure
```{r, fig.height=15, fig.width=12}
# assemble main figure
plot_grid(
  dhs_baseline_panel,
  abc_baseline_panel,
  plot_grid(prc_pub, performance_pub, rel_widths = c(0.8, 1), labels = c("c", "d")),
  plot_grid(corr_plot, crispr_corr, nrow = 1, rel_widths = c(1, 0.8), labels = c("e", "f")), 
  labels = c("a", "b", NA, NA),
  rel_heights = c(1, 1, 0.8, 0.9),
  ncol = 1
)

# save figure to pdf file
ggsave(filename = "results/manuscript/plots/extended_data_fig2_crispr_benchmarking.pdf",
       height = 15, width = 12)
```

***

## Session information
This document was produced using following packages:
```{r sessionInfo}
sessionInfo()
```
